Parallel Monte Carlo Synthetic Acceleration Methods for Discrete Transport Problems

### 2013-08-16 doc/dissertation/presentation/defense.tex
adding a publications slide to the defense talk

  \begin{itemize}
  \item Monte Carlo Synthetic Acceleration Methods
    \bigskip
  \item Application to Neutron Transport
    \bigskip
  \item Application to Fluid Flow
    \bigskip
  \item Parallelization of MCSA
  \end{itemize}

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{Research Outline}
  \begin{itemize}
  \item Development of a linear solver for discrete systems leveraging
    Monte Carlo Synthetic Acceleration
    \medskip
    \begin{itemize}
    \item Application to neutron transport
    \item Research is required to explore general solver development
    \item Performance is of concern
    \end{itemize}
    \bigskip
  \item Development of a nonlinear solver for discrete systems
    leveraging Monte Carlo Synthetic Acceleration
    \begin{itemize}
    \item Application to fluid flow
    \item Potential memory benefits
    \item Convergence of the linear model is of concern
    \end{itemize}
    \bigskip
  \item Parallelization of Monte Carlo Synthetic Acceleration
    \medskip
    \begin{itemize}
    \item Parallel strategies taken from modern reactor physics
      methods
    \item Research is required to explore varying parallel strategies
    \item Scalability is of concern
    \end{itemize}
  \end{itemize}
\end{frame}

@@

  \begin{itemize}
  \item \textbf{Monte Carlo Synthetic Acceleration Methods}
    \bigskip
  \item Application to Neutron Transport
    \bigskip
  \item Application to Fluid Flow
    \bigskip
  \item Parallelization of MCSA
  \end{itemize}

@@

  \begin{itemize}
  \item Preliminary development during January of 2013
    \bigskip
  \item General asynchronous MSOD MCSA implementation
    \begin{itemize}
    \item Forward and adjoint Monte Carlo with method of expected
      values
    \item Parallel row matrix/vector interface
    \item General fixed point iteration strategy
    \item Explicit algebraic preconditioner suite
    \end{itemize}
    \medskip
  \item Implemented in C++
    \bigskip
  \item Heavy use of the Trilinos scientific computing libraries
    \bigskip
  \item Open-source BSD 3-clause license
    \bigskip
  \item https://github.com/sslattery/MCLS
  \end{itemize}

@@

  \begin{itemize}
  \item Monte Carlo Synthetic Acceleration Methods
    \bigskip
  \item \textbf{Application to Neutron Transport}
    \bigskip
  \item Application to Fluid Flow
    \bigskip
  \item Parallelization of MCSA
  \end{itemize}

@@

      \begin{figure}[htpb!]
        \begin{center}
          \includegraphics[width=1.9in]{solver_iters.pdf}
        \end{center}
      \end{figure}

@@

  \begin{itemize}
  \item Monte Carlo Synthetic Acceleration Methods
    \bigskip
  \item Application to Neutron Transport
    \bigskip
  \item \textbf{Application to Fluid Flow}
    \bigskip
  \item Parallelization of MCSA
  \end{itemize}

@@

  \begin{table}[h!]
    \begin{center}
      \begin{tabular}{lcc}\hline\hline
        \multicolumn{1}{l}{Benchmark}& 
        \multicolumn{1}{c}{Newton-Krylov}&
        \multicolumn{1}{c}{FANM}\\
        \hline
        Convection, Ra=\sn{1}{3} & 5 & 5 \\
        Convection, Ra=\sn{1}{4} & 7 & 7 \\
        Convection, Ra=\sn{1}{5} & 9 & 10 \\
        Convection, Ra=\sn{1}{6} & 11 & 11 \\
        \hline
        Lid Driven, Re=100 & 6 & 6 \\
        Lid Driven, Re=300 & 9 & 9 \\
        Lid Driven, Re=500 & 11 & 11 \\
        Lid Driven, Re=700 & 14 & 10 \\
        \hline
        Backward Step, Re=200 & 10 & 9 \\
        Backward Step, Re=300 & 15 & 14 \\
        Backward Step, Re=400 & 10 & 10 \\
        Backward Step, Re=500 & 19 & 20 \\
        %%
        \hline\hline
      \end{tabular}`
    \end{center}
    \caption{\textbf{Navier-Stokes benchmark comparison for nonlinear
        iterations.} \textit{Over all benchmarks, FANM performed better
        in terms of nonlinear iterations for 1 more case than the
        Newton-Krylov method.}}
    \label{tab:benchmark_nonlinear_comparison}

@@

  \begin{table}[h!]
    \begin{center}
      \begin{tabular}{lcc}\hline\hline
        \multicolumn{1}{l}{Benchmark}& 
        \multicolumn{1}{c}{Newton-Krylov}&
        \multicolumn{1}{c}{FANM}\\
        \hline
        Convection, Ra=\sn{1}{3} & 32 & 18 \\
        Convection, Ra=\sn{1}{4} & 23 & 17 \\
        Convection, Ra=\sn{1}{5} & 25 & 20 \\
        Convection, Ra=\sn{1}{6} & 39 & 25 \\
        \hline
        Lid Driven, Re=100 & 27 & 42 \\
        Lid Driven, Re=300 & 35 & 52 \\
        Lid Driven, Re=500 & 41 & 56 \\
        Lid Driven, Re=700 & 21 & 14 \\
        \hline
        Backward Step, Re=200 & 24 & 13 \\
        Backward Step, Re=300 & 23 & 17 \\
        Backward Step, Re=400 & 18 & 12 \\
        Backward Step, Re=500 & 30 & 52 \\
        %%
        \hline\hline
      \end{tabular}
    \end{center}
    \caption{\textbf{Navier-Stokes benchmark comparison for total linear
        solver iterations.} \textit{Over all benchmarks, FANM performed
        better in terms of linear solver iterations for twice as many
        cases as the Newton-Krylov method.}}
    \label{tab:benchmark_linear_comparison}
  \end{table}

@@

  \begin{itemize}
  \item Monte Carlo Synthetic Acceleration Methods
    \bigskip
  \item Application to Neutron Transport
    \bigskip
  \item Application to Fluid Flow
    \bigskip
  \item \textbf{Parallelization of MCSA}
  \end{itemize}

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{Parallel Scaling}

@@

@@

  \small{
    \begin{itemize}
    \item MCSA can solve the asymmetric system generated by the $SP_N$
      equations
      \medskip
    \item Light water reactor problems are difficult to solve with MCSA as
      they have large spectral radii due to the neutron scattering in the
      moderator
      \medskip
    \item Advanced algebraic preconditioning strategies were applied to
      the $SP_N$ equations to obtain convergence with ILUT chosen for
      subsequent investigations
      \medskip
    \item MCSA was verified to produce the same flux distribution and
      k-eigenvalue for the fuel assembly as production Krylov methods
      \medskip
    \item MCSA was observed to converge in fewer iterations per eigenvalue
      iteration than GMRES for the fuel assembly criticality problem and
      more than Bi-CGStab using the same preconditioning
    \end{itemize}
  }

@@

  \small{
    \begin{itemize}
    \item Forward-Automated Newton-MCSA (FANM) has been developed
      \medskip
    \item The FANM method has been verified to produce the same solutions
      as a production Newton-Krylov method for three difficult benchmark
      problems
      \medskip
    \item The FANM method has better iterative performance than the
      Newton-Krylov method for convection dominated problems, converging
      in fewer linear solver iterations with the same preconditioning for
      high and low Rayleigh numbers
      \medskip
    \item The spectral radius convergence restriction on MCSA was observed
      to be a significant hindrance by preventing solutions to forced flow
      problems at high Reynolds numbers
      \medskip
    \item More Monte Carlo histories at every FANM iteration can reduce
      the number of linear and nonlinear iterations required to converge
      the problem
    \end{itemize}
  }

@@

  \small{
    \begin{itemize}
    \item The multiple-set overlapping-domain (MSOD) parallel algorithm
      for particle transport has been adapted to parallelize MCSA
      \medskip
    \item MCSA scales favorably compared to production Krylov methods for
      both strong and weak scaling cases
      \medskip
    \item Overlap in small quantities can provide parallel efficiency
      boosts of a few percent in strong scaling cases but is
      ineffective in weak scaling cases
      \medskip
    \item Multiple sets offers a means to reduce time to solution by
      solving multiple copies of the original problem and combining the
      solutions using superposition
      \medskip
    \item MCSA is most efficiently used in parallel as a stochastic
      realization of an additive Schwarz method
    \end{itemize}
  }

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{Discussion}

>>>>

  \begin{itemize}
  \item Monte Carlo Synthetic Acceleration Methods
    \bigskip
  \item Application to Neutron Transport
    \bigskip
  \item Application to Fluid Flow
    \bigskip
  \item Parallelization of MCSA
    \bigskip
  \item Summary
  \end{itemize}

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{Research Outline}
  \begin{itemize}
  \item Development of a linear solver for discrete systems leveraging
    Monte Carlo Synthetic Acceleration
    \medskip
    \begin{itemize}
    \item Application to neutron transport
    \item Research is required to explore general solver development
    \item Performance is of concern
    \end{itemize}
    \bigskip
  \item Development of a nonlinear solver for discrete systems
    leveraging Monte Carlo Synthetic Acceleration
    \begin{itemize}
    \item Monte Carlo is a more natural fit
    \item Application to fluid flow
    \item Convergence of the linear model is of concern
    \end{itemize}
    \bigskip
  \item Parallelization of Monte Carlo Synthetic Acceleration
    \medskip
    \begin{itemize}
    \item Parallel strategies taken from modern reactor physics
      methods
    \item Research is required to explore varying parallel strategies
    \item Scalability is of concern
    \end{itemize}
  \end{itemize}
\end{frame}

@@

  \begin{itemize}
  \item \textbf{Monte Carlo Synthetic Acceleration Methods}
    \bigskip
  \item Application to Neutron Transport
    \bigskip
  \item Application to Fluid Flow
    \bigskip
  \item Parallelization of MCSA
    \bigskip
  \item Summary
  \end{itemize}

@@

  \begin{itemize}
  \item Designed to be easily incorporated with production physics
    codes
    \bigskip
  \item General asynchronous MSOD MCSA implementation
    \begin{itemize}
    \item Forward and adjoint Monte Carlo with method of expected
      values
    \item Parallel row matrix/vector interface
    \item General fixed point iteration strategy
    \item Explicit algebraic preconditioner suite
    \end{itemize}
    \medskip
  \item Implemented in C++
    \bigskip
  \item Heavy use of the Trilinos scientific computing libraries
    \bigskip
  \item Open-source BSD 3-clause license
    \bigskip
  \item https://github.com/sslattery/MCLS
  \end{itemize}

@@

  \begin{itemize}
  \item Monte Carlo Synthetic Acceleration Methods
    \bigskip
  \item \textbf{Application to Neutron Transport}
    \bigskip
  \item Application to Fluid Flow
    \bigskip
  \item Parallelization of MCSA
    \bigskip
  \item Summary
  \end{itemize}

@@

      {\tiny
      \begin{table}[h!]
        \begin{center}
          \begin{tabular}{lccc}\hline\hline
            \multicolumn{1}{l}{Solver}&
            \multicolumn{1}{c}{1 Group}&
            \multicolumn{1}{c}{2 Groups}&
            \multicolumn{1}{c}{4 Groups}\\
            \hline
            BiCGStab-ILUT & 11.6 & 11.6 & 12.4 \\ 
            GMRES-ILUT & 18.1 & 17.9 & 18.9 \\
            MCSA-ILUT-R-C & 14.6 & 15.4 & 17.6 \\
            MCSA-ILUT-MR-C & 16.0 & 17.1 & 23.7 \\
            MCSA-ILUT-R-EV & 18.3 & 19.4 & 16.8 \\
            MCSA-ILUT-MR-EV & 19.6 & 22.4 & 17.5 \\
            Richardson-ILUT & 60.9 & 60.4 & 63.4 \\
            %%
            \hline\hline
          \end{tabular}
        \end{center}
        \caption{Average number of linear solver iterations per
          eigenvalue iteration.}
      \end{table}
      }

@@

  \begin{itemize}
  \item Monte Carlo Synthetic Acceleration Methods
    \bigskip
  \item Application to Neutron Transport
    \bigskip
  \item \textbf{Application to Fluid Flow}
    \bigskip
  \item Parallelization of MCSA
    \bigskip
  \item Summary
  \end{itemize}

@@

  \begin{table}[h!]
    \begin{center}
      \begin{tabular}{lccc}\hline\hline
        \multicolumn{1}{l}{Benchmark}& 
        \multicolumn{1}{c}{NK}&
        \multicolumn{1}{c}{FANM}&
        \multicolumn{1}{c}{NR}\\
        \hline
        Convection, Ra=\sn{1}{3} & 5 & 5 & 5\\
        Convection, Ra=\sn{1}{4} & 7 & 7 & 7\\
        Convection, Ra=\sn{1}{5} & 9 & 10 & 9\\
        Convection, Ra=\sn{1}{6} & 11 & 11 & 11\\
        \hline
        Lid Driven, Re=100 & 6 & 6 & 7\\
        Lid Driven, Re=300 & 9 & 9 & 9\\
        Lid Driven, Re=500 & 11 & 11 & 11\\
        Lid Driven, Re=700 & 14 & 10 & 12\\
        \hline
        Backward Step, Re=200 & 10 & 9 & 10\\
        Backward Step, Re=300 & 15 & 14 & 15\\
        Backward Step, Re=400 & 10 & 10 & 10\\
        Backward Step, Re=500 & 19 & 20 & 21\\
        %%
        \hline\hline
      \end{tabular}`
    \end{center}
    \caption{\textbf{Navier-Stokes benchmark comparison for nonlinear
        iterations.} \textit{Over all benchmarks, FANM performed better
        in terms of nonlinear iterations for 1 more case than the
        Newton-Krylov method.}}
    \label{tab:benchmark_nonlinear_comparison}

@@

  \begin{table}[h!]
    \begin{center}
      \begin{tabular}{lccc}\hline\hline
        \multicolumn{1}{l}{Benchmark}& 
        \multicolumn{1}{c}{GMRES}&
        \multicolumn{1}{c}{MCSA}&
        \multicolumn{1}{c}{Richardson}\\
        \hline
        Convection, Ra=\sn{1}{3} & 32 & 18 & 38\\
        Convection, Ra=\sn{1}{4} & 23 & 17 & 34\\
        Convection, Ra=\sn{1}{5} & 25 & 20 & 34\\
        Convection, Ra=\sn{1}{6} & 39 & 25 & 48\\
        \hline
        Lid Driven, Re=100 & 27 & 42 & 151\\
        Lid Driven, Re=300 & 35 & 52 & 133\\
        Lid Driven, Re=500 & 41 & 56 & 154\\
        Lid Driven, Re=700 & 21 & 14 & 32\\
        \hline
        Backward Step, Re=200 & 24 & 13 & 21\\
        Backward Step, Re=300 & 23 & 17 & 21\\
        Backward Step, Re=400 & 18 & 12 & 14\\
        Backward Step, Re=500 & 30 & 52 & 98\\
        %%
        \hline\hline
      \end{tabular}
    \end{center}
    \caption{\textbf{Navier-Stokes benchmark comparison for total linear
        solver iterations.} \textit{Over all benchmarks, FANM performed
        better in terms of linear solver iterations for twice as many
        cases as the Newton-Krylov method.}}
    \label{tab:benchmark_linear_comparison}
  \end{table}

@@

  \begin{itemize}
  \item Monte Carlo Synthetic Acceleration Methods
    \bigskip
  \item Application to Neutron Transport
    \bigskip
  \item Application to Fluid Flow
    \bigskip
  \item \textbf{Parallelization of MCSA}
    \bigskip
  \item Summary
  \end{itemize}

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{Leadership-Class Parallel Scaling Studies}

@@

  \begin{itemize}
  \item Monte Carlo Synthetic Acceleration Methods
    \bigskip
  \item Application to Neutron Transport
    \bigskip
  \item Application to Fluid Flow
    \bigskip
  \item Parallelization of MCSA
    \bigskip
  \item \textbf{Summary}
  \end{itemize}

@@

  \small{
    \begin{itemize}
    \item MCSA has been incorporated into the Exnihilo neutronics
      production code base developed at Oak Ridge National Laboratory
      \medskip
    \item MCSA can solve the asymmetric system generated by the $SP_N$
      equations
      \medskip
    \item Light water reactor problems are difficult to solve with MCSA as
      they have large spectral radii due to the neutron scattering in the
      moderator
      \medskip
    \item Advanced algebraic preconditioning strategies were applied to
      the $SP_N$ equations to obtain convergence with ILUT chosen for
      subsequent investigations
      \medskip
    \item MCSA was observed to converge in fewer iterations per eigenvalue
      iteration than GMRES for the fuel assembly criticality problem and
      more than Bi-CGStab using the same preconditioning
    \end{itemize}
  }

@@

  \small{
    \begin{itemize}
    \item Forward-Automated Newton-MCSA (FANM) has been developed
      \medskip
    \item The FANM method has been incorporated into the Drekar
      multiphysics production code base developed at Sandia National
      Laboratories
      \medskip
    \item The FANM method has better iterative performance than the
      Newton-Krylov method for convection dominated problems, converging
      in fewer linear solver iterations with the same preconditioning for
      high and low Rayleigh numbers
      \medskip
    \item The spectral radius convergence restriction on MCSA was observed
      to be a significant hindrance by preventing solutions to forced flow
      problems at high Reynolds numbers
      \medskip
    \item More Monte Carlo histories at every FANM iteration can reduce
      the number of linear and nonlinear iterations required to converge
      the problem
    \end{itemize}
  }

@@

  \small{
    \begin{itemize}
    \item The multiple-set overlapping-domain (MSOD) parallel
      algorithm for domain decomposed particle transport has been
      adapted to parallelize MCSA
      \medskip
    \item MCSA scales favorably compared to production Krylov methods for
      both strong and weak scaling cases
      \medskip
    \item Overlap in small quantities can provide parallel efficiency
      boosts of a few percent in strong scaling cases but is
      ineffective in weak scaling cases
      \medskip
    \item Multiple sets offers a means to reduce time to solution by
      solving multiple copies of the original problem and combining the
      solutions using superposition
      \medskip
    \item MCSA is most efficiently used in parallel as a stochastic
      realization of an additive Schwarz method
    \end{itemize}
  }

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{Publications}

  \small{
    \begin{enumerate}
    \item S.R. Slattery, T.M. Evans, P.P.H. Wilson, \textbf{A
      Multiple-Set Overlapping-Domain Decomposed Monte Carlo Synthetic
      Acceleration Method for Linear Systems}, \textit{Joint
      International Conference on Supercomputing in Nuclear
      Applications and Monte Carlo 2013 (SNA+MC 2013), Paris, France,
      October 27-31, 2013. Accepted for oral presentation.}
      \medskip
    \item T.M. Evans, S.W. Mosher, S.R. Slattery, S.P. Hamilton,
      \textbf{A Monte Carlo Synthetic-Acceleration Method for Solving
        the Thermal Radiation Diffusion Equation}, \textit{Journal of
        Computational Physics, Submitted.}
      \medskip
    \item S.R. Slattery, T.M. Evans, P.P.H. Wilson, \textbf{A Spectral
      Analysis of the Domain Decomposed Monte Carlo Method for Linear
      Systems}, \textit{International Conference on Mathematics and
      Computational Methods Applied to Nuclear Science \& Engineering
      (M\&C 2013), American Nuclear Society, Sun Valley, ID, May 5-9,
      2013.}
    \end{enumerate}
  }
\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Thank You}

  \begin{itemize}
  \item Paul
  \item Tom
  \item Roger
  \item CASL, ORNL, Sandia, OLCF
\end{itemize}
